\large \textbf{Leveraging Historical Ties between Cognitive Science and Computer Science to Guide Innovative Use of Computers in Education}

### 2016-02-02 paper.tex
another wave of edits and additions
still need to rework discussion, future work, and add a conclusion.

\titleformat*{\section}{\normalsize\bfseries} % Makes section titles 12 pt font
\titleformat{\section}{\normalsize\bfseries}{}{}{}[]
\titleformat{\subsection}{\bfseries}{}{}{}[]

@@

% conflict between teaching programming syntax, teaching ``how to program'', teaching high-level concepts like abstractions and how to ``do'' problem solving.
\section{Coding for Everyone}
In the past few years, there has been increasing interest in a
movement to get more people, regardless of background, to learn to
program, in fact President Obama recently made a statement calling on
all children to have the opportunity to learn about computer
science\autocite{whitehouse_computer_2016}. Sites such as code.org
promote CS education opportunities for all citing statistics about
STEM jobs and arguing that in the 21st century knowledge about
computer science is foundational\autocite{code.org_every_2016}. Many
institutions of higher education either have programming requirements
for non-computer-science majors, or have been expanding programming to
non-majors\autocite{rich_cs1_2004,forte_motivation_2005,guzdial_design_2005}. At
the authors of this paper's institution, all first year general
engineering students are required to complete two 6-week rotations of
an introductory programming curriculum as part of their first-year
engineering course.

Juxtaposed with the observation that technology is changing rapidly,
and the half-life of technical skills is
decreasing\autocite{nae_educating_2004} it would seem teaching
programming is a questionable use of class time: the language and
paradigm students are taught will likely not be the same they are
asked to use in their first job post-graduation. However, if we
acknowledge that learning to program is more than simply learning to
write code, and in particular if we draw on work from the early era of
computing which closely linked programming to problem solving and
cognitive skills, we can argue that regardless of language or
paradigm, learning to program may actually be a way to learn to
problem solve, reason about and organize knowledge, and conceptualize
processes.

While there are some recent studies which investigate programming
education as a means of teaching problem-solving, there is also a
large body of research that primarily focuses on teaching students the
details of how to code. We argue that in most cases these details are
less important than centering programming education around
problem-solving skills and if educators hope to achieve this goal then
all aspects of introductory programming courses, from the course
objectives, to the curriculum used, to the assessments administered,
must be chosen intentionally with this goal in mind. In this paper we
will present a brief summary of current published work on introductory programming education, then note a potential lack that, 
then offer a potential framework to

@@

\section{Methods}
[identify learning sciences sources]

Assuming that there would be more comprehensive research of
introductory programming in the computer science community we began
with targeted searchs of the ACM digital library for `introductory
programming'. Once we found key papers that seemed to indicate a call
to action
\autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
or a survey of practices
\autocite[e.g.][]{pears_survey_2007-1,robins_learning_2003} we used
the `cited by' list to find additional relevant papers.

@@

\section{Current state of research}
%%TODO: add citations
The bulk of literature about teaching introductory programming (CS1)
revolves around what paradigms are ``best'' (e.g. procedural vs. object
oriented), which language is ``best'', often times with regard to a
particular paradigm (e.g. OOP with Python vs. Java), and best teaching
practices (e.g. use of “pair programming” environments). However, the
\emph{reasons} for picking one paradigm over another tend to be about
industry demand, rather than intentionally choosing a paradigm to
support the learning objectives of the course.
%% TODO: reminder to connect to learning science, e.g. this behavior
%% is not uncommong, for instance flipped classrooms is a ``hot
%% topic'' now and often times we see people implementing a ``flipped
%% classroom'' without necessarily considering if that pedagogical
%% approach makes the most sense for a given course.

\section{Overall direction}
A \citeyear{robins_learning_2003} study by
\citeauthor{robins_learning_2003} identifies four trends in research
into programming education: 1) distinguishing between novice and
expert programmers with emphasis on deficiencies of novices, 2) the
distinction between knowledge and strategies, 3) the distinction
between program comprehension and generation, and 4) a comparison of
OOP to procedural styles of programming\autocite{robins_learning_2003}.

The \textcite{mccracken_multinational_2001} study, which concluded that
regardless of nation or institution, most students do not know how to
program by the end of their introductory programming courses, sparked
much conversation and push to reform introductory programming in the
next decade. Ten years after the
\citeauthor{mccracken_multinational_2001} study,
\textcite{lister_ten_2011} reflected on the response and lamented that
while much of the criticism that the problems used to test student
knowledge were too difficult, no one had repeated the study with
problems they deemed to be more reasonable. 
%% TODO: connect mccracken to engineering ed about missconceptions
%% (CHEER chapter and Streveler paper)

Thus while the \citeauthor{mccracken_multinational_2001} study
validated the experiences of many introductory programming instructors
it did not result in critical change to address the problems. Rather,
what we have seen since the 1990s is been an increasing interest in
where object oriented programming should be placed within the
curriculum with debates between objects-first vs procedural-fist but
little question of why object oriented design should be included at
all other than it continues to be a popular paradigm in industry.

Much fewer, but still some, authors have drawn on cognitive sciences
to determine how indroductory programming courses should be
taught. \textcite{muller_pattern_2005} propose a pattern oriented
approach that places emphasis on recognizing and learning to work with
algorithmic patterns over syntax and design paradigms.

There remains a desire that introductory programming courses should
teach problem-solving skills, but few papers present evidence that the
implementation of this goal is informed by cognitive science.

@@

A large number of papers surveyed contain discussions of course
objectives, pedagogical tools used, and assessments given, but rarely
are these three components discussed together.

@@

Assessment results can help inform changes to course objectives and
instructional methods\autocite{felder_designing_2003}. Similarly, as
instructional methods improve, the objectives may be modified to aim
for higher levels of understanding. Being intentional about these
interactions is crucial to an effective learning system; if the
assessment addresses lower levels of learning than the objectives
specify, ``the system will be driven by backwash from testing, not by
the curriculum.''\autocite[][p. 350]{biggs_enhancing_1996}. For
example, if an objective of a programming course is that learners will
be able to transfer high level concepts from programming to other
domains, but assessment only consists of checking that programs
function conforms to data input/output specifications,
teaching/learning activities would tend to center around those that
promote correct input/output behavior and exclude activities that
might have a better chance at promoting higher level understanding and
transfer\autocite{biggs_enhancing_1996}.

\section{Case Study}
\subsection{Object Oriented Programming}
As noted, teaching OOP is a common area of interest in the literature.

It should be strongly noted that we are not advocating for OOP as a
``best'' teaching paradigm. Rather, we want to point out that from what
has been reported it seems that it is often not taught
effectively. What educators seem to miss is that OOP defines an
entirely different paradigm for problem solving than is used in
procedural programming\autocite{kolling_problem_1999}. As should not
be surprising there are some problems for which OOP provides a good
set of tools to solve in an efficient, elegant manner, namely those
with solutions taht naturally map to real-world
objects\autocite{raymond_art_2003}. Different problems may be better
solved using a procedural, or functional
paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
our goal should be to help students to analyze problem structures and
make informed decisions regarding what paradigm would be best for a
particular problem. What is not clear from the literature is whether
or not this consideration is made when designing problems for a
particular course. Indeed, in some cases there is evidence that OOP
concepts are added to a course ``on top of'' older content, without
necessarily changing the problem sets (citation needed).

Alan Kay originally conceived of object oriented programming as as one
way to think about complex systems\autocite{kay_history_1996}. The
underlying metaphore of Kay's conception of OOP was that of the
biological cell: each object is designed to accomplish a specific set
of tasks where the complexity of these tasks is encapsulated within
the object. Objects interact with one another by sending and receiving
messages and systems of these objects are built to solve complex
problems\autocite{kay_history_1996}.

@@

\section{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. In an
attempt to address the concern that the two primary languages, C++ and
Java, used to teach OOP in introductory courses, may be too complex
for novice students, \textcite{goldwasser_teaching_2008} chooses
Python to be the language of choice for introductory programming
claiming that its relatively simple syntax allow new students to
better engage with the OOP concepts rather than getting bogged down by
details of the language.

>>>>

\titleformat*{\section}{\normalsize\bfseries} % Makes section titles 12 pt font
\titleformat{\section}{\normalsize\bfseries}{}{}{}[]
\titleformat{\subsection}{\bfseries}{}{}{}[]
\titleformat{\subsubsection}{\bfseries}{}{}{}[]

@@

% conflict between teaching programming syntax, teaching ``how to program'', teaching high-level concepts like abstractions and how to ``do'' problem solving.
\section{Coding for Everyone}
In the past few years, there has been increasing interest to get more
people, regardless of background, to learn to program. In fact
President Obama recently made a statement calling on all children to
have the opportunity to learn about computer
science\autocite{whitehouse_computer_2016}. Sites such as code.org
promote CS education opportunities for all, citing statistics about
STEM jobs and arguing that in the 21st century knowledge about
computer science is foundational\autocite{code.org_every_2016}. Many
institutions of higher education either have programming requirements
for non-computer-science majors, or have been expanding programming to
non-majors\autocite{rich_cs1_2004,forte_motivation_2005,guzdial_design_2005}. At
the authors of this paper's institution, all first year general
engineering students are required to complete an introductory
programming module as part of their first-year engineering course.

Juxtaposed with the observation that technology is changing rapidly,
and the half-life of technical skills is
decreasing\autocite{nae_educating_2004}, it would seem that teaching
programming is a questionable use of class time: the language and
paradigm students are taught will likely not be the same they are
asked to use in their first job post-graduation. However, if we
acknowledge that learning to program is more than simply learning to
write code, and in particular if we draw on work from the early era of
computing which closely linked programming to problem solving and
cognitive skills, we can argue that regardless of language or
paradigm, learning to program may actually be a way to learn to
problem solve, organize knowledge, and conceptualize processes.

While there are some recent studies which investigate programming
education as a means of teaching problem-solving, there is also a
large body of research that primarily focuses on teaching students the
details of how to code. We argue that in most cases these details are
less important than centering programming education around
problem-solving skills. If we make learning problem-solving skills a
goal of introductory programming courses, then all aspects of the
course, from the course objectives, to the curriculum used, to the
assessments administered, must be chosen intentionally with this goal
in mind.

In this paper we will present a brief summary of current published
work on introductory programming education. We will note that much of
the current work lacks a tight integration of course objectives,
pedagogical strategies, and assessments, and will offer a potential
framework to help think about both the design and evaluation of
introductory programming courses.

@@

\section{Methods}
Assuming that there would be more comprehensive research of
introductory programming in the computer science community we began
with targeted searchs of the ACM digital library for `introductory
programming'. Once we found key papers that seemed to indicate a call
to action
\autocite[e.g.][]{mccracken_multinational_2001,kolling_problem_1999},
or a survey of practices
\autocite[e.g.][]{pears_survey_2007-1,robins_learning_2003} we used
the `cited by' list to find additional relevant papers.

We needed a framework to consistently evaluate and starting from
\citeauthor{jamieson_creating_2009}'s
(\citeyear{jamieson_creating_2009}) to ground educational research
activities in how people learn we identified the concept of
\empha{constructive alignment} as a promising one. Following the
references in \citeauthor{jamieson_creating_2009}'s report as well as
conducting targeted literature searches for ``constructive alignment''
we were able to find descriptions of this
framework\autocite{jamieson_creating_2009,pellegrino_rethinking_2006,felder_designing_2003,biggs_enhancing_1996}.

@@

\section{Current state of research}
The bulk of literature about teaching introductory programming (CS1)
revolves around what paradigms are ``best'' (e.g. procedural
vs. object oriented
\autocite{bailie_objects_2003,moritz_objectsfirst_2005,reges_back_2006}),
which language is ``best'', often times with regard to a particular
paradigm (e.g. OOP with Python
vs. Java\autocite{goldwasser_teaching_2008,koulouri_teaching_2014}),
and best teaching practices (e.g. use of “pair programming”
environments\autocite{williams_support_2001,mcdowell_effects_2002,thomas_code_2003,chong_social_2007,salleh_empirical_2011}). However,
the \emph{reasons} for picking one paradigm over another tend to be
about industry demand, rather than intentionally choosing a paradigm
to support the learning objectives of the course. This behavior,
making a popular practice part of a learning environment without
critically analyzing whether or not is a good fit is by no means
exclusive to programming education. For example, flipped
classrooms\autocite{bishop_flipped_2013}, in which students are asked
to watch recorded lectures at home and use class time for working on
activities, has recently been a hot topic in higher education and
there have been a continuous number of examples of well-intentioned
practitioners flipping their classroom without adjusting course
objectives, content and activities to work under the flipped-classroom
model\autocite{desantis_students_2015}. Likewise, anyone can teach
elements of object oriented programming in their introductory
programming courses, but unless critical aspects of the course are
adjusted: background reading, rethinking what the ``fundamental
concepts'' are, the types of problems assigned, and methods of
assessment, the incorporation of OOP will likely create confusion for
students and frustration for instructor\autocite{kolling_problem_1999}.

%% TODO: reminder to connect to learning science, e.g. this behavior
%% is not uncommong, for instance flipped classrooms is a ``hot
%% topic'' now and often times we see people implementing a ``flipped
%% classroom'' without necessarily considering if that pedagogical
%% approach makes the most sense for a given course.

\section{Overall direction}
A \citeyear{robins_learning_2003} study by
\citeauthor{robins_learning_2003} identifies four trends in research
into programming education: 1) distinguishing between novice and
expert programmers with emphasis on deficiencies of novices, 2) the
distinction between knowledge and strategies, 3) the distinction
between program comprehension and generation, and 4) a comparison of
OOP to procedural styles of
programming\autocite{robins_learning_2003}. They found that the
biggest challenge for novice programmers was not understanding the
basic concepts, but knowing how to apply them. This finding is
consistent understanding of how we
learn\autocite{bransford_how_2000,ambrose_chapter_2010}, and with
studies of expert/novice behavior in other domains, such a
statics\autocite{litzinger_cognitive_2010}. In both cases the
suggestion is that helping students learn strategies that work is
crucial to helping them become effective novices, i.e. devote
instructional time to problem-solving strategies and how to apply
them.

The \citeyear{mccracken_multinational_2001}
\citeauthor{mccracken_multinational_2001}
study\autocite{mccracken_multinational_2001}, which concluded that
regardless of nation or institution, most students do not know how to
program by the end of their introductory programming courses, sparked
much conversation and push to reform introductory programming in the
next decade. This phenomenon is by no means exclusive to introductory
programming: disconnect between what we think we are teaching and what
students learn has been observed across
engineering\autocite{streveler_learning_2008,flynn_engineering_2014}. Possible
ways of addressing this disconnect include recognizing differences in
student's prior knowledge, helping students see connections between
abstract concepts, and relating difficult concepts to ones students
already know\autocite{streveler_learning_2008,slotta_helping_2006}.
%% TODO: connect mccracken to engineering ed about missconceptions
%% (CHEER chapter and Streveler paper)

While the \citeauthor{mccracken_multinational_2001} study validated the
experiences of many introductory programming instructors it did not
result in critical change to address the problems. Rather, what we
have seen since the 1990s is been an increasing interest in where
object oriented programming should be placed within the curriculum
with debates between objects-first vs procedural-fist but little
question of why object oriented design should be included at all other
than it continues to be a popular paradigm in industry.

Much fewer, but still some, authors have drawn on cognitive sciences
to determine how indroductory programming courses should be
taught. \textcite{muller_pattern_2005} propose a pattern oriented
approach that places emphasis on recognizing and learning to work with
algorithmic patterns over syntax and design paradigms. This approach
closely aligns with understanding of differences in expert/novice
behavior by helping students identify patterns of solutions and
learning how and when to apply them to particular problems.

There remains a desire that introductory programming courses should
teach problem-solving skills, though only a few papers present
evidence that the implementation of this goal is informed by cognitive
science\autocite{kramarski_cognitivemetacognitive_1997,reed_incorporating_1998,muller_supporting_2008}.

@@

Finally, a \citeyear{koulouri_teaching_2014} study by
\citeauthor{koulouri_teaching_2014} conducted a quantitative analysis
of different approaches to teaching introductory programming by
varying three aspects: the language used, providing formative
feedback, and introducing explicit problem solving
training\autocite{koulouri_teaching_2014}. They found that choosing a
language with a simpler syntax (Python over Java), and introducing
explicit problem-solving training improved student learning, while
providing more formative assessment did not seem to help. They
speculate that the reason for the latter finding could be that
students do not spend time looking at formative feedback, or do not
understand what is given.

We wish to build upon the past reviews of the field by suggesting an
approach to designing learning systems that may more consistently
address some of the shortcomings observed, and bring understanding to
the successes. A large number of papers surveyed contain discussions
of course objectives, pedagogical tools used, and assessments given,
but rarely are these three components discussed together. We know from
the engineering education literature that an effective learning
system depends on the alignment of these three
components\autocite{jamieson_creating_2009,pellegrino_rethinking_2006}.

@@

Assessment results can help inform changes to course objectives and
instructional methods\autocite{felder_designing_2003}. Similarly, as
instructional methods improve, the objectives may be modified to aim
for higher levels of understanding. Being intentional about these
interactions is crucial to an effective learning system; if the
assessment addresses lower levels of learning than the objectives
specify, ``the system will be driven by backwash from testing, not by
the curriculum.''\autocite[][p. 350]{biggs_enhancing_1996}. For
example, if an objective of a programming course is that learners will
develop problem solving skills, but assessment only consist of test of
programming concepts and syntax, teaching/learning activities would
tend to center around those that promote understanding of these
concepts and exclude activities that might have a better chance at
promoting general problem solving\autocite{biggs_enhancing_1996}.

\section{Case Study}
As noted, teaching OOP is a common area of interest in the literature.

It should be strongly noted that we are not advocating for OOP as a
``best'' teaching paradigm. Rather, we want to point out that from
what has been reported it seems that it is often not taught
effectively. What educators seem to miss is that OOP defines an
entirely different paradigm for problem solving than is used in
procedural programming\autocite{kolling_problem_1999}. As should not
be surprising there are some problems for which OOP provides a good
set of tools to solve in an efficient, elegant manner, namely those
with solutions taht naturally map to real-world
objects\autocite{raymond_art_2003}. Different problems may be better
solved using a procedural, or functional
paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
our goal should be to help students to analyze problem structures and
make informed decisions regarding what paradigm would be best for a
particular problem. What is not clear from the literature is whether
or not this consideration is made when designing problems for a
particular course. Indeed, in some cases there is evidence that OOP
concepts are added to a course ``on top of'' older content, without
necessarily changing the problem
sets\autocite{kolling_problem_1999,vilner_fundamental_2007}.

\subsection{Object Oriented Programming}

Alan Kay originally conceived of object oriented programming as
\emph{one} way to think about complex
systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
conception of OOP was that of the biological cell: each object is
designed to accomplish a specific set of tasks where the complexity of
these tasks is encapsulated within the object. Objects interact with
one another by sending and receiving messages, and systems of these
objects are assembled to solve complex
problems\autocite{kay_history_1996}.

@@

In several reports of attempts to incorporate OOP into introductory
programming courses, practitioners worried that time devoted to OOP
concepts would necessarily reduce time spent on the ``basics of
programming'', usually defined as conditional statements, loops, and
in some cases pointers. This concern is misplaced, under an OOP
paradigm the ``basics of programming'' are not the same as when
working from a procedural paradigm. OOP defines a completely different
set of tools for problem solving, and it is these tools and concepts
that become the
``basics''\autocite{kolling_problem_1999,kay_history_1996}.

For instance, if taking a patterns-based approach to teaching
programming one could use either a procedural paradigm or an OOP
paradigm, but the fundamental patterns would be quite different
depending on the paradigm\autocite{wallingford_first_1996}.

Ultimately, if our goal is to teach problem solving in general, we
would be doing a grave disservice to our students to leave with the
impression that there is only one paradigm, whether it be OOP,
procedural, functional, or something not yet invented. As previously
mentioned, OOP and procedural paradigms are comprised of fundamentally
different concepts for problem solving, and so again is the functional
paradigm. Unsurprisingly, there is no paradigm that is the ``best''
fit for every problem. Leaving students with the believe that only one
paradigm exists makes it very difficult for them to learn how to
intentionally select and apply the ``best'' paradigm to solve a
particular problem. \citeauthor{kolling_problem_1999} argues that
performing a paradigm shift from procedural style to OOP is difficult,
and so if OOP is a desired outcome then this should be taught
\emph{first}.

\section{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. The
explaination for this can be found in cognitive load theory (CLT),
which posits that learners have a finite capacity for their short-term
memory, and for learning to occur, all aspects of the task must fit
into this finite space. While some contributors to cognitive load are
necessary for the task at hand, others are
not\autocite{paas_cognitive_2003}. If, for example the goal is to
teach problem-solving techniques there will be cognitive load
associated with thinking about different strategies one could
use. Learning a programming language's syntax rules also adds to the
cognitive load, but it does not aid in the process of learning
problem-solving techniques. Thus, it is in the educators best interest
to minimize cognitive load associated with features that do not
directly help with the learning objectives of the course.

In an attempt to address the concern that the two primary languages,
C++ and Java, used to teach OOP in introductory courses, may be too
complex for novice students, \textcite{goldwasser_teaching_2008}
chooses Python to be the language of choice for introductory
programming claiming that its relatively simple syntax allow new
students to better engage with the OOP concepts rather than getting
bogged down by details of the language.

\section{Problem choice}
As previously noted, OOP provides a set of tools that naturally lend
themselves to solving certain types of problems, but not others. An
educator teaching OOP should carefully select problems and challenges
that make sense to be solved with an OOP paradigm. Of course,
provoiding illustrative examples of problems that are \emph{not} best
solved with an OOP approach would be useful provided there is also
time for discussion of \emph{why} OOP might not be suited for certain
problems and brief mention of alternative approaches.
